---
title: "Assignment 3: Dangerous Foul Balls"
author: "Phil Graff"
date: "10/5/2020"
output: html_document
---

In the summer of 2019, I planned to take my partner, father, and four-year old daughter out to a baseball game. Earlier that spring, I was horrified to watch a video clip of a batter hitting a foul ball into the stands where it struck a two-year old child (Albert Almora batting against Houston Astros). I knew my priority would be to find seats for my family behind protective netting.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
# Import foul-balls.csv downloaded from Five Thirty-Eight's GitHub
# https://github.com/fivethirtyeight/data/blob/master/foul-balls/foul-balls.csv
foulballs <- read.csv("foul-balls.csv")
foulcounts <- table(foulballs$used_zone)
foulpct <- foulcounts / nrow(foulballs)
```


```{r}
# Assign colors
library(RColorBrewer)

# derive number of color bins
bins <- ceiling(max(foulpct*10))

# Create palette for the number of bins
palette <- brewer.pal(n = bins, name = "Greens")

# Assign palette value to each stadium region
hex_vector <- palette[ceiling(foulpct*10)]

# Pass palette vector into D3 script to render field and color stands
library(r2d3)
r2d3(data=hex_vector, script = "baseballd3.js")

```

